---
title: "10+ Guidelines for Better Tables in R"
description: |
  Make tables people ACTUALLY want to read.
author:
  - name: Thomas Mock
    url: https://twitter.com/thomas_mock
base_url: https://themockup.blog
date: 09-04-2020
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gt)
tuesdata <- tidytuesdayR::tt_load(2020, "36")
```

```{r yield-data-hide, echo=FALSE}
country_sel <- c("China", "India", "United States", "Indonesia", "Mexico", "Pakistan")

yield_data_wide <- tuesdata$key_crop_yields %>% 
  janitor::clean_names() %>% 
  rename_with(~str_remove(., "_tonnes_per_hectare")) %>% 
  select(entity:beans, -code) %>% 
  pivot_longer(cols = wheat:beans, names_to = "crop", values_to = "yield") %>% 
  rename(Country = entity) %>% 
  filter(crop %in% c("potatoes", "maize"), 
         year %in% c(2014:2016),
         Country %in% country_sel) %>% 
  pivot_wider(names_from = year, values_from = yield)
```

## Basic `gt` Table

You can create a table by passing in data to `gt()`, and the idea is that you progressively add layers/changes to the `gt` table via the pipe.

```{r}
# This works!
# gt(yield_data_wide)

# pipe also works!
yield_data_wide %>% 
  gt()
```

```{r}
yield_data_wide %>% 
  head() %>% 
  group_by(Country) %>% 
  gt(rowname_col = "crop")
```

```{r}
yield_data_wide %>% 
  head() %>% 
  gt(
    groupname_col = "crop",
    rowname_col = "Country"
  )
```

```{r}
yield_data_wide %>% 
  mutate(crop = str_to_title(crop)) %>% 
  group_by(crop) %>% 
  gt(
    rowname_col = "Country"
  ) %>% 
  fmt_number(
    columns = 2:5,
    decimals = 2
  ) %>% 
  summary_rows(
    groups = TRUE,
    columns = vars(`2014`, `2015`, `2016`),
    fns = list(
      Average = ~mean(.x),
      SD = ~sd(.x)
    )
  )
```

```{r}
yield_data_wide %>% 
  head() %>% 
  gt(
    groupname_col = "crop",
    rowname_col = "Country"
  ) %>% 
  tab_spanner(
    label = "Yield in Tonnes/Hectare",
    columns = 2:5
  )
  
```

```{r}
yield_data_wide %>% 
  head() %>% 
  gt(
    groupname_col = "crop",
    rowname_col = "Country"
  ) %>% 
  tab_header(
    title = md("**Crop Yields between 2014 and 2016**"),
    subtitle = md("*Countries limited to Asia*")
  )
```

```{r}
yield_data_wide %>% 
  head() %>% 
  gt(
    groupname_col = "crop",
    rowname_col = "Country"
  ) %>% 
  tab_header(
    title = "Crop Yields between 2014 and 2016",
    subtitle = "Conutries limited to Asia"
  ) %>% 
  tab_options(
    heading.subtitle.font.size = 12,
    heading.align = "left",
    table.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(3)
  )
```

```{r}
yield_data_wide %>% 
  head() %>% 
  gt() %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "black", alpha = 0.2),
      cell_borders(
        side = c("left", "right"),
        color = "black",
        weight = px(2)
      )
    ),
    locations = cells_body(
      columns = vars(crop)
    )
  ) %>% 
  tab_style(
    style = cell_text(font = google_font("IBM Plex Mono")),
    locations = cells_body(columns = 2:5)
  ) %>% 
  tab_style(
    style = cell_text(color = "#F7AE7E", style = "italic"),
    locations = cells_body(
      columns = 3:5,
      rows = Country == "China"
    )
  )
```

```{r}
yield_data_wide %>% 
  head() %>% 
  gt(
    groupname_col = "crop",
    rowname_col = "Country"
  ) %>% 
  data_color(
    columns = vars(`2014`, `2015`, `2016`),
    colors = scales::col_numeric(c("white", "pink", "red"), domain = NULL)
    )
```

## Data prep

```{r}
country_sel <- c("China", "India", "United States", "Indonesia", "Mexico", "Pakistan")

yield_data <- 
  tuesdata$key_crop_yields %>% 
  janitor::clean_names() %>% 
  rename_with(~str_remove(., "_tonnes_per_hectare")) %>% 
  select(entity:beans, -code) %>% 
  pivot_longer(cols = wheat:beans, names_to = "crop", values_to = "yield") %>% 
  rename(Country = entity)
```

## Rule 1: Offset the heads from the body

```{r}
# data prep
potato_data <- yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% c(2013:2016)) %>% 
  pivot_wider(names_from = year, values_from = "yield")
```

```{r}
# poor example
potato_tb <- potato_data %>% 
  gt() %>% 
  cols_hide(vars(crop)) %>% 
  opt_table_lines(extent = "none") %>% 
  fmt_number(
    columns = 3:6,
    decimals = 2
  )

potato_tb
```

```{r}
# improved example
potato_tb %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>% 
  opt_table_lines(extent = "default") %>% 
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.bottom.color = "black",
    table_body.hlines.color = "white",
    table.border.bottom.color = "white"
  ) %>% 
  tab_source_note(
    md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br />**Inspiration**: @jschwabish")
  )
```
## Rule 2: Use subtle dividers rather than heavy gridlines

```{r}
# data perp

rule2_data <- 
  yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% c(2007:2016)) %>% 
  select(-crop) %>% 
  pivot_wider(names_from = year, values_from = yield) %>% 
  rowwise() %>% 
  mutate(
    avg_07_11 = mean(`2007`:`2011`),
    .before = `2012`
  ) %>% 
  mutate(
    avg_12_16 = mean(`2012`:`2016`)
  ) %>% 
  ungroup()
```

```{r}
# Poor example
rule2_data %>% 
  gt(rowname_col = "Country") %>% 
  cols_label(
    avg_07_11 = "Avg.",
    avg_12_16 = "Avg."
  ) %>% 
  cols_width(
    1 ~ px(125)
  ) %>% 
  fmt_number(
    columns = 2:last_col()
  ) %>% 
  tab_style(
    style = cell_borders(
      side = "all",
      color = "grey",
      weight = px(1),
      style = "solid"
    ),
    locations = list(
      cells_body(everything()),
      cells_column_labels(everything())
    )
  ) %>% 
  grand_summary_rows(
    columns = 2:last_col(),
    fns = list(
      "Avg" = ~ mean(.x)
    ),
    formatter = fmt_number
  ) %>% 
  tab_style(
    style = cell_text(
      align = "right",
      color = "grey"
    ),
    locations = cells_stub()
  ) %>% 
  tab_options(
    grand_summary_row.background.color = "lightblue",
    alig
  )
```

```{r}
rule2_data %>% 
  add_row(
    rule2_data %>% 
      summarise(
        across(where(is.double), list(Average = mean), .names = "{col}")
      ) %>% 
      mutate(Country = "Average")
  ) %>% 
  gt() %>% 
  cols_label(
    avg_07_11 = "Avg.",
    avg_12_16 = "Avg."
  ) %>% 
  fmt_number(
    columns = 2:last_col()
  ) %>% 
  tab_style(
    style = cell_fill(
      color = "lightgrey",
      alpha = 0.4 
    ),
    locations = list(
      cells_body(columns = vars(avg_07_11, avg_12_16)),
      cells_column_labels(columns = vars(avg_07_11, avg_12_16))
    )
  ) %>% 
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "black",
      weight = px(3)
    ),
    locations = cells_body(
      columns = everything(),
      rows = Country == "Average"
    )
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>% 
  tab_options(
    column_labels.border.top.color = "black",
    column_labels.border.top.width = px(3),
    column_labels.border.bottom.color = "black"
  )
```

## Rule 3: Right-align numbers and heads

```{r}
# Prep data
rule3_data <- 
  yield_data %>% 
  filter(Country == "United States", year %in% 2016) %>% 
  mutate(crop = str_to_title(crop)) %>% 
  pivot_wider(names_from = "year", values_from = yield) %>% 
  arrange(crop) %>% 
  select(-Country, Crop = crop)
```

```{r}
rule3_data %>% 
  mutate(
    `Center align` = `2016`,
    `Right align` = `2016`
  ) %>% 
  rename(`Left align` = `2016`) %>% 
  gt() %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    location = cells_column_labels(everything())
  ) %>% 
  fmt_number(
    columns = 2:4
  ) %>% 
  cols_align(align = "left", columns = "Left align") %>% 
  cols_align(align = "center", columns = "Center align") %>% 
  cols_align(align = "right", columns = "Right align") %>% 
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.bottom.color = "black",
    table_body.hlines.color = "white"
  )
```

## Rule 4: Left-align Text and Heads

## Rule 5: Select the appropriate level of precision

## Rule 6: Guide your reader with space between rows and columns

```{r}
rule6_data <- yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% 2014:2016) %>% 
  pivot_wider(names_from = "year", values_from = "yield") %>% 
  select(-crop)

rule6_tb <- rule6_data %>% 
  add_row(
    rule6_data %>% 
      summarise(across(where(is.double), list(Average = mean), .names = "{col}")) %>% 
      mutate(Country = "Average")
  ) %>% 
  gt() %>% 
  fmt_number(
    columns = 2:4,
    decimals = 2
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>% 
  tab_style(
    style = cell_borders(sides = "top", color = "black", weight = px(2)),
    locations = cells_body(columns = everything(), rows = Country == "Average")
  ) %>% 
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3)
  ) %>% 
  cols_width(vars(Country) ~ px(125), 2:4 ~ px(75))
```

## Rule 7: Remove unit repetition

```{r}
rule6_tb %>% 
  fmt_percent(
    columns = 2:4,
    rows = 1,
    scale_values = FALSE
  ) %>% 
  cols_align(align = "left", columns = 2:4)
```

```{r}
rule6_tb %>% 
  tab_spanner(
    label = "% Yield of Total",
    columns = everything()
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold", align = "right", size = "xx-small"),
    locations = cells_column_spanners("% Yield of Total")
  )
```

```{r}
rule6_tb %>% 
  tab_footnote(
    footnote = md("**% Yield of Total**"),
    locations = cells_column_labels(columns = 2:4)
  )
```

## Rule 8: Highlight outliers

```{r}
rule8_data <- yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% 2009:2017) %>% 
  group_by(Country) %>% 
  mutate(pct_change = (yield / lag(yield) - 1)*100) %>% 
  ungroup() %>% 
  filter(between(year, 2010, 2016)) %>% 
  select(Country, year, pct_change) %>% 
  pivot_wider(names_from = "year", values_from = pct_change)
```

```{r}
rule8_data %>% 
  gt() %>% 
  fmt_number(2:last_col()) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>% 
  cols_width(
    vars(Country) ~ px(125),
    2:last_col() ~ px(75)
  ) %>% 
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.bottom.color = "black"
  ) %>% 
  tab_style(
    style = list(cell_fill(scales::alpha("red", 0.7)), cell_text(color = "white", weight = "bold")),
    locations = list(
      cells_body(columns = 2, rows = `2010` < 0),
      cells_body(columns = 3, rows = `2011` < 0),
      cells_body(columns = 4, rows = `2012` < 0),
      cells_body(columns = 5, rows = `2013` < 0),
      cells_body(columns = 6, rows = `2014` < 0),
      cells_body(columns = 7, rows = `2015` < 0),
      cells_body(columns = 8, rows = `2016` < 0)
    )
  )
```

## Rule 9: Group similar data and increase white space

## Rule 10: Add visualizations when appropriate

```{r}
rule10_data <- yield_data %>% 
  filter(year %in% c(2013, 2017), crop == "potatoes", Country %in% country_sel) %>% 
  pivot_wider(names_from = year, values_from = yield)
```

```{r}
plot_spark <- function(data){
  data %>% 
    mutate(
      yield_start = if_else(year == 2013, yield, NA_real_),
      yield_end = if_else(year == 2017, yield, NA_real_)
    ) %>% 
    tidyr::fill(yield_start, yield_end, .direction = "downup") %>% 
    mutate(color = if_else(yield_end - yield_start < 0, "red", "blue")) %>% 
    ggplot(aes(x = year, y = yield, color = color)) +
    geom_line(size = 15) +
    theme_void() +
    scale_color_identity() +
    theme(legend.position = "none")
}

yield_plots <- yield_data %>% 
  filter(
    year %in% 2013:2017,
    crop == "potatoes",
    Country %in% country_sel
  ) %>% 
  nest(yields = c(year, yield)) %>% 
  mutate(plot = map(yields, plot_spark))

rule10_data %>% 
  mutate(ggplot = NA) %>% 
  select(-crop) %>% 
  gt() %>% 
  text_transform(
    locations = cells_body(vars(ggplot)),
    fn = function(x){
      map(yield_plots$plot, ggplot_image, height = px(15), aspect_ratio = 4)
    }
  ) %>% 
  cols_width(vars(ggplot) ~ px(100)) %>% 
  cols_label(
    ggplot = "2013-2017"
  ) %>% 
  fmt_number(2:3) %>% 
  tab_spanner(
    label = "Potato Yield in Tonnes/Hectare",
    columns = c(2, 3)
  ) %>% 
  tab_style(
    style = cell_text(color = "black", weight = "bold"),
    locations = list(
      cells_column_labels(everything()),
      cells_column_spanners(everything())
    )
  ) %>% 
  tab_options(
    row_group.border.top.width = px(3),
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "black",
    table_body.hlines.color = "white",
    table.border.top.color = "white",
    table.border.top.width = px(3),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(2)
  )
```
